Synthesizing Quotient Lenses

### 2017-10-12 papers/QuotientLensSynth/main.tex

\begin{abstract}
{\em A Quotient lens} is an expression which encapsulates a bidirectional
transformation whereby each of the two transformations that are part of the lens
have the same behaviour on data that are equivalent to each other modulo some
equivalence relation. Quotient lenses save the programmer the effort of writing
two expressions instead of one, give guarantess on how the forward and backward
transformations interact with each other, and also allow the programmer to
ignore differences in the data which are not intended to affect the
behaviour of the transformations.

However, quotient lenses are sometimes difficult to program, firstly because the
programmer has to give a specification for the data as well as give a separate
specification for the equivalence relations defined on the data. Secondly, even
after giving both of these specifications, the programmer also has to manually
write the underlying lens which maps between the respective equivalence
classes, a task which is itself challenging.

In this paper, we address both of these challenges. First, we propose the
language of {\em Quotient Regular Expressions} or QREs, which enables the
programmer to explicitly define a broad class of equivalence relations
directly on regular languages using a compact and convenient notation, and hence
negates the need to specify the data and the equivalence relation on the data
separately. We then define a class of {\em QRE lenses}, a class of quotient
lenses which are typed using QREs, and show how to {\em synthesize} QRE lenses
from their QRE types and a set of input-output examples. Finally, we
demonstrate the applicability of this approach by synthesizing QRE lenses
between several real-world data formats from the {\tt data.gov} database.
\end{abstract}

@@

\noindent stored in the \bibtex format to their equivalent representation 
\begin{verbatim}
%0 Book
%T Regular Algebra and Finite Machines
%A Conway, J. H.
%F conway
\end{verbatim}

@@

Unfortunately, quotient lenses are sometimes difficult to program in Boomerang,
firstly because the programmer has to give a specification for the data as well
as give a separate specification for the equivalence relations defined on the
data. For example, suppose that the programmer wants to program the \bibtex to
EndNote transformation in such a way that the program allows for any of two
representations of the authors name (e.g. ``Conway, J. H." and ``Conway J. H"),
ignores the order in which the author and title fields occur in each record,
and allows for arbitrary amounts of whitespace in between the author and title
fields. These rules allow for these two \bibtex citation records to be
considered the same,

@@

In Boomerang, equivalence relations are expressed using {\em canonizers}
which are functions that map each element of a regular language to a
canonical representation of that element. Therefore, in order for the programmer
to give a specification for the \bibtex records, then the programmer
has to write a regular expression which matches all possible representations of
\bibtex records, and then program a set of canonizers which does the job of
quotienting the data. This often makes the code verbose and error-prone.

@@

To address this issue, we propose the language of {\em Quotient Regular
Expressions} or QREs, which enables the programmer to explicitly define a broad
class of equivalence relations directly on regular languages using a compact
and convenient notation, and hence negates the need to specify the data and the
equivalence relation on the data separately. For example, a QRE which
specifies the equivalence relation on the \bibtex citation records is as
follows:

@@

However, even though QREs enable the programmer to specify the data as well as
the equivalence relation on the data in the same expression, the programmer
still has to manually program the underlying lens which maps the equivalence
classes of the \bibtex records to the equivalence classes of the EndNote
records

@@

to derive a quotient lens that does the \bibtex to EndNote transformation.

@@

For example, suppose that the programmer wishes to write a regular expression
which describes \bibtex citation records such as:
\begin{verbatim}
@Book {conway,
  Author = {Conway, J. H.},
  Title = {Regular Algebra and Finite Machines},
}
\end{verbatim}

Each record may be represented in many different ways. Firstly, the author's
name can have the last name followed by a comma, and then the
first and second names separated by a space, or the last name, first name and
second name separated by spaces in between them. For example, each of these
\bibtex records is equivalent to the other:
\begin{verbatim}
@Book {conway,
  Author = {Conway, J. H.},
  Title = {Regular Algebra and Finite Machines},
}
\end{verbatim}
\begin{verbatim}
@Book {conway,
  Author = {Conway J. H.},
  Title = {Regular Algebra and Finite Machines},
}
\end{verbatim}

@@

  Author = {Conway, J. H.},
}
\end{verbatim}
In the example above, a \bibtex record may have different
representations for three different reasons. The first reason
is that the author's name can have the last name followed by a comma and then
the first and second names separated by a space, or the last name, first name and
second name separated by spaces in between them. The $\squash{R}{R'}{f}$ QRE
enables the programmer to choose between the representations described by $R$
and $R$', provided that they can provide a function $f$ that maps data stored in
the first format to data stored in the second format. In this example we have
the two regular expressions

@@

The second reason that a record may have two different representations is that
the author and title fields may occur in a different order, and the third
reason is that the author and title fields may have arbitrary occurences of
whitespace characters in between them. The $R \mapsto s$ and $\perm{q_1,
\ldots, q_n}{q}$ QREs enable the programmer to handle each of these cases:
$$\perm{\re{AUTHOR}, \re{TITLE}}{((``," \cdot \re{WHITESPACE}^*) \mapsto \string
``, \backslash n")}$$
This QRE enables the programmer to consider any permutation of the author and
title fields interspersed with a comma and arbitrary whitespace as equivalent.
The $((``," \cdot \re{WHITESPACE}^*) \mapsto \string ``, \backslash n")$
subexpression is itself a QRE that denotes that all string $s$ which consists
of a comma $``,"$ followed by any number of whitespace should be considered
equivalent, with the canonical representative being the string $\string ``,
\backslash n"$. The canonical representation of \bibtex records is that in
which the author field occurs the title field with a comma then a newline
character separating the two.

The regular combinators for QREs i.e. $\cdot$ (concatenation), $|$ (union) and
$*$ (iteration) behave like the regular combinators for regular languages. For
instance, if the QRE for matching a \bibtex \; record is named
$\mathtt{RECORD}$, then the QRE for matching two records is $\mathtt{RECORD}
\cdot \mathtt{RECORD}$, the QRE for matching one or two records is $\mathtt{RECORD}
\sep (\mathtt{RECORD} \cdot \mathtt{RECORD})$ and the QRE for matching any
number of QRE records in $\mathtt{RECORD}^*$.

@@

\subsection{Semantics of QREs}
\subsubsection{Preliminaries}
Let $q, q'$ be QREs. When applying the regular combinators to QREs, we require
that if a string $s$ matches any of the regular expressions $W(q) \cdot W(q')$,
$W(q) \sep W(q')$, $W(q)^*$, $K(q) \cdot K(q')$,
$K(q) \sep K(q')$, $K(q)^*$, then $s$ matches that regular expression in
only one way. This unambiguity condition is called {\em strong unambiguity},
and is necessary, firstly because it ensures that the canonizing function of a
QRE is well-defined. For example if $\mathtt{RECORD}$ is the QRE for a \bibtex
record in our \bibtex to and $W(\mathtt{ENDNOTE} \cdot \mathtt{ENDNOTE})$ is not
unambiguous, that is there is a string $s$ such that $s = s_1 \cdot s_2 =
{s_1}' \cdot {s_2}'$ with $s_1, s_2, {s_1}', {s_2}' \in W(\mathtt{ENDNOTE} \cdot
\mathtt{ENDNOTE})$, then the canonizer $f$ for $\mathtt{ENDNOTE} \cdot
\mathtt{ENDNOTE}$ may not be well defined since we are not guaranteed that
$f(s_1) \cdot f(s_2) = f({s_1}') \cdot f({s_2}')$. We also require that the
regular combinators applied to the kernels are unambiguous since the underlying
lenses will end up operating on the kernels, and these lenses impose the same
restrictions for similar reasons.

>>>>

\begin{abstract}

  {\em Quotient lenses} express bidirectional transformations between data
  formats, where the inputs to the transformation are considered modulo some
  equivalence relation.  Such lenses save the programmer from having to write
  two expressions and satisfy strong ``lens laws'' that ensure that the forward
  and backward transformations interact well.  The use of quotients allows the
  programmer to ignore ``inessential'' differences in the data, such as the
  amount whitespace or the ordering of fields.

  However, quotient lenses are often difficult to program.  The programmer has
  to give a specification for the data on each side of the lens and a separate
  specification for the equivalence relations defined on the data. The
  programmer must also manually implement a lens program that maps between the
  specifications while respecting the equivalence relations---this is itself a
  challenging task thanks to the restrictions on how lenses can be combined
  while still satisfying the lens laws.

This paper addresses both of these challenges. First, we propose the
language of {\em Quotient Regular Expressions}, or QREs, which 
define a broad class of equivalence relations
directly on regular languages using a compact and convenient notation. This
eliminates the need to specify the data and the equivalence relation on the data
separately. We then define {\em QRE lenses}, a class of quotient
lenses that are typed using QREs, and show how to {\em synthesize} QRE lenses
from their QRE types and a set of input-output examples. Finally, we
demonstrate the applicability of this approach by synthesizing QRE lenses
between several real-world data formats from the {\tt data.gov} database.
\end{abstract}

@@

\noindent stored in the \bibtex{} format to their equivalent representation 
\begin{verbatim}
%0 Book
%T Regular Algebra and Finite Machines
%A Conway, J. H.
%F conway
\end{verbatim}

@@

Unfortunately, quotient lenses are sometimes difficult to program in Boomerang,
firstly because the programmer has to give a specification for the data as well
as give a separate specification for the equivalence relations defined on the
data. For example, suppose that the programmer wants to program the \bibtex{} to
EndNote transformation in such a way that the program allows for any of two
representations of the authors name (e.g. ``Conway, J. H." and ``Conway J. H"),
ignores the order in which the author and title fields occur in each record,
and allows for arbitrary amounts of whitespace in between the author and title
fields. These rules allow for these two \bibtex{} citation records to be
considered the same,

@@

In Boomerang, equivalence relations are expressed using {\em canonizers}
which are functions that map each element of a regular language to a
canonical representation of that element. Therefore, in order for the programmer
to give a specification for the \bibtex{} records, then the programmer
has to write a regular expression which matches all possible representations of
\bibtex{} records, and then program a set of canonizers which does the job of
quotienting the data. This often makes the code verbose and error-prone.

@@

To address this issue, we propose the language of {\em Quotient Regular
Expressions} or QREs, which enables the programmer to explicitly define a broad
class of equivalence relations directly on regular languages using a compact
and convenient notation, and hence negates the need to specify the data and the
equivalence relation on the data separately. For example, a QRE which
specifies the equivalence relation on the \bibtex{} citation records is as
follows:

@@

However, even though QREs enable the programmer to specify the data as well as
the equivalence relation on the data in the same expression, the programmer
still has to manually program the underlying lens which maps the equivalence
classes of the \bibtex{} records to the equivalence classes of the EndNote
records

@@

to derive a quotient lens that does the \bibtex{} to EndNote transformation.

@@

For example, suppose that the programmer wishes to write a regular expression
which describes \bibtex{} citation records such as:
\begin{verbatim}
@Book {conway,
  Author = {Conway, J. H.},
  Title = {Regular Algebra and Finite Machines},
}
\end{verbatim}

Each record may be represented in many different ways. Firstly, the author's
name can have the last name followed by a comma, and then the
first and second names separated by a space, or the last name, first name and
second name separated by spaces in between them. For example, each of these
\bibtex{} records is equivalent to the other:
\begin{verbatim}
@Book {conway,
  Author = {Conway, J. H.},
  Title = {Regular Algebra and Finite Machines},
}
\end{verbatim}
\begin{verbatim}
@Book {conway,
  Author = {Conway J. H.},
  Title = {Regular Algebra and Finite Machines},
}
\end{verbatim}

@@

  Author = {Conway, J. H.},
}
\end{verbatim}
In the example above, a \bibtex{} record may have different
representations for three different reasons. The first reason
is that the author's name can have the last name followed by a comma and then
the first and second names separated by a space, or the last name, first name and
second name separated by spaces in between them. The $\squash{R}{R'}{f}$ QRE
enables the programmer to choose between the representations described by $R$
and $R$', provided that they can provide a function $f$ that maps data stored in
the first format to data stored in the second format. In this example we have
the two regular expressions

@@

The second reason that a record may have two different representations is that
the author and title fields may occur in a different order, and the third
reason is that the author and title fields may have arbitrary occurences of
whitespace characters in between them. The $R \mapsto s$ and $\perm{q_1,
\ldots, q_n}{q}$ QREs enable the programmer to handle each of these cases:
$$\perm{\re{AUTHOR}, \re{TITLE}}{((``," \cdot \re{WHITESPACE}^*) \mapsto \string
``, \backslash n")}$$
This QRE enables the programmer to consider any permutation of the author and
title fields interspersed with a comma and arbitrary whitespace as equivalent.
The $((``," \cdot \re{WHITESPACE}^*) \mapsto \string ``, \backslash n")$
subexpression is itself a QRE that denotes that all string $s$ which consists
of a comma $``,"$ followed by any number of whitespace should be considered
equivalent, with the canonical representative being the string $\string ``,
\backslash n"$. The canonical representation of \bibtex{} records is that in
which the author field occurs the title field with a comma then a newline
character separating the two.

The regular combinators for QREs i.e. $\cdot$ (concatenation), $|$ (union) and
$*$ (iteration) behave like the regular combinators for regular languages. For
instance, if the QRE for matching a \bibtex{} \; record is named
$\mathtt{RECORD}$, then the QRE for matching two records is $\mathtt{RECORD}
\cdot \mathtt{RECORD}$, the QRE for matching one or two records is $\mathtt{RECORD}
\sep (\mathtt{RECORD} \cdot \mathtt{RECORD})$ and the QRE for matching any
number of QRE records in $\mathtt{RECORD}^*$.

@@

\subsection{Semantics of QREs}
\subsubsection{Preliminaries}
Let $q, q'$ be QREs. When applying the regular combinators to QREs, we require
that if a string $s$ matches any of the regular expressions $W(q) \cdot W(q')$,
$W(q) \sep W(q')$, $W(q)^*$, $K(q) \cdot K(q')$,
$K(q) \sep K(q')$, $K(q)^*$, then $s$ matches that regular expression in
only one way. This unambiguity condition is called {\em strong unambiguity},
and is necessary, firstly because it ensures that the canonizing function of a
QRE is well-defined. For example if $\mathtt{RECORD}$ is the QRE for a \bibtex
record in our \bibtex{} to and $W(\mathtt{ENDNOTE} \cdot \mathtt{ENDNOTE})$ is not
unambiguous, that is there is a string $s$ such that $s = s_1 \cdot s_2 =
{s_1}' \cdot {s_2}'$ with $s_1, s_2, {s_1}', {s_2}' \in W(\mathtt{ENDNOTE} \cdot
\mathtt{ENDNOTE})$, then the canonizer $f$ for $\mathtt{ENDNOTE} \cdot
\mathtt{ENDNOTE}$ may not be well defined since we are not guaranteed that
$f(s_1) \cdot f(s_2) = f({s_1}') \cdot f({s_2}')$. We also require that the
regular combinators applied to the kernels are unambiguous since the underlying
lenses will end up operating on the kernels, and these lenses impose the same
restrictions for similar reasons.
